---
title: "Assignment 2: Coding Basics"
author: "Emma Childs"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff=45), tidy=TRUE)
```

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document.
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Sakai.

## Basics Day 1

1.  Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name.

2.  Compute the mean and median of this sequence.

3.  Ask R to determine whether the mean is greater than the median.

4.  Insert comments in your code to describe what you are doing.

```{r}
#1. 
firsthomeworkseq<-seq(1,100,4)
firsthomeworkseq
#I named the sequence firsthomeworkseq and then put the inputs as 1 and 100, which are the start and end points. Then I added a comma and the number 4, as that is what I wanted wanted the sequence to count by.

#2. 
mean(firsthomeworkseq)
median(firsthomeworkseq)
#I used the mean and median functions in R to calculate these values, and used firsthomeworkseq as my input. I got the same number - 49 - for the mean and median.

#3. 
mean(firsthomeworkseq)>median(firsthomeworkseq)
#I simply ran a true/false type of statement to determine if the mean was greater than the median for this input sequence. It comes back as false because they are equal to each other. 
```

## Basics Day 2

5.  Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6.  Label each vector with a comment on what type of vector it is.

7.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

8.  Label the columns of your data frame with informative titles.

```{r}
names <- c("Mason", "Hildegard", "Gertrude", "Fannie Lou")
scores <- c("49", "93", "87", "97")
passing <- c("FALSE", "TRUE", "TRUE", "TRUE")

#For types of vectors, names is characters, scores is numeric, passing is also comprised of characters

Student.Test.Scores <- data.frame("Names" = names, "Test Score" = scores, "Passing" = passing)
```

9.  QUESTION: How is this data frame different from a matrix?

> Answer: I believe matrices can only contain one form of data, and this data frame contains both character and numeric data, setting it apart. 

10. Create a function with an if/else statement. Your function should take a **vector** of test scores and print (not return) whether a given test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. 

11. Apply your function to the vector with test scores that you created in number 5.

```{r}
Student.Fail.Function <- function(x){
  x <- ifelse(x<50, "FAIL", "PASS")
 print(x)
}
Student.Fail.Function(100)
Student.Fail.Function(40)
```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer:I found that "ifelse" worked, but I'm not exactly sure why, to be totally honest. I was trying to follow along how we did it on the class video, but I'm going to ask for more help this week! Thanks so much. 
