---
title: "Assignment 5: Data Visualization"
author: "Emma Childs"
date: "Spring 2023"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1.  Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Upload the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 
getwd()

install.packages("tidyverse")
install.packages("lubridate")
install.packages("here")
install.packages("cowplot")

library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
#installing and loading packages 

NTL.Chem.Nutrients.PeterPaul <- read.csv(file=here("./Data/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)
NEON.litter <- read.csv(file=here("./Data/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)
#reading in files

#2 
NEON.litter$collectDate <- ymd(NEON.litter$collectDate)

NTL.Chem.Nutrients.PeterPaul$sampledate <- ymd(NTL.Chem.Nutrients.PeterPaul$sampledate)
#Used 'lubridate' to make sure dates are being read as date formats
```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3
my_theme <- theme_classic(base_size = 12) + 
  theme(
    plot.background = element_rect(
      color = 'black',
      fill = 'lemonchiffon'
    ),
    line = element_line(
      color='purple',
      linewidth =2
    ),
    legend.background = element_rect(
      color='grey',
      fill = 'white'
    ),
    legend.title = element_text(
      color='black'
        ))
    theme_set(my_theme)
    
    #defining and setting custom theme after playing around with colors
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 
NTL.LTER <- NTL.Chem.Nutrients.PeterPaul %>%
  ggplot(
    aes(x=po4,
      y=tp_ug,
      color=lakename)
    ) + 
  geom_point(size=0.5, alpha=0.5) +
    labs(
      title = "Phosphorus and Phosphate Levels for Peter & Paul Lakes",
      x="Phosphate (po4)",
      y="Phosphorus (ug)", 
      color="Lakename"
    ) +
  geom_smooth(
    method = 'lm',
    se = FALSE,
    color = 'black') +
    xlim(0,50)

print(NTL.LTER)
#Plotted total phosphorus by phosphate and added best fit line. 
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tip: R has a build in variable called `month.abb` that returns a list of months;see <https://r-lang.com/month-abb-in-r-with-example>

```{r Create boxplots}
#5 
#Part a.
box_temps <- NTL.Chem.Nutrients.PeterPaul %>%
  ggplot(
    aes(x=factor(month,level=1:12,labels = month.abb),
      y=temperature_C,
      color=lakename)
  ) + 
  geom_boxplot(size=0.5, alpha=0.5) +
    labs(
      title = "Temperatures for Peter & Paul Lakes",
      x="Month",
      y="Temperature (C)", 
      color="Lakename"
    ) 

print(box_temps)

#Part b.
box_TP <- NTL.Chem.Nutrients.PeterPaul %>%
  ggplot(
    aes(x=factor(month,level=1:12,labels = month.abb),
      y=tp_ug,
      color=lakename)
  ) + 
  geom_boxplot(size=0.5, alpha=0.5) +
    labs(
      title = "Total Phosphorus for Peter & Paul Lakes",
      x="Month",
      y="Total Phosphorus (ug)", 
      color="Lakename"
    ) 

print(box_TP)

#Part c.
box_TN <- NTL.Chem.Nutrients.PeterPaul %>%
  ggplot(
    aes(x=factor(month,level=1:12,labels = month.abb),
      y=tn_ug,
      color=lakename)
  ) + 
  geom_boxplot(size=0.5, alpha=0.5) +
    labs(
      title = "Total Nitrogen for Peter & Paul Lakes",
      x="Month",
      y="Total Nitogren (ug)", 
      color="Lakename"
    ) 

print(box_TN)

#Coded all three box plots and ran them successfully, I think!

#Part d/cowplot.
#install.packages("cowplot") - already done
library(cowplot)

plot_grid(box_temps, box_TP, box_TN, labels = c('Temperature (C)', 'Total Phosporus', 'Total Nitrogen'), label_size = 12)

#p1 <- ggplot(box_temps, aes(x=month, y=temperature_C)) + geom_point()
#p2 <- ggplot(box_TP, aes(x=month, y=tp_ug)) + geom_point()
#p3 <- ggplot(box_TN, aes(x=month, y=tn_ug)) + geom_point()
#This last section may not be correct, but I was playing around with different methods and tried this one. 

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: It is a little difficult to say from just looking at the graphs, but it seems like the temperatures are relatively similar between lakes. Peter Lake seems to have higher phosphorus, with highest values in the summer months, and pretty similar with total nitrogen as well in comparison to Paul Lake. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6
#Plotting "Needles" functional group by date and separating with color
litter_subset <- NEON.litter %>% 
  filter(functionalGroup=='Needles') %>% 
  ggplot(aes(
      x=collectDate,
      y=dryMass,
      color=nlcdClass)
    ) + 
  geom_point(size=0.5,alpha=0.5) +
    labs(
    title="Dry Mass of Needle Litter",
    x="Date",
    y="Dry Mass",
    color="NLCD Class"
    )

print(litter_subset)

#7
#Plotting dry mass again, but faceted by NLCD class
litter.subset.faceted <- NEON.litter %>%
  filter(functionalGroup=='Needles') %>% 
  ggplot(
     aes(x=collectDate, 
         y=dryMass,)) +
  geom_point(size=0.5,alpha=0.5) +
  facet_wrap(vars(nlcdClass)) +
    labs(
    title="Dry Mass of Needle Litter",
    x="Date",
    y="Dry Mass"
    )

print(litter.subset.faceted)

```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I find number 6 more effective because I think having all the data on one graph with different colors allows for better comparision between NLCD classes than three separate graphs. 